---
title: "Chpater_2, Workflow: basics"
format: gfm
editor: visual
---

# Coding Basics
```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```
```{r}
x <- 3 * 4
print(x)
```
You can combine multiple elements into a vector with c():

```{r}
primes <- c(2, 3, 5, 7, 11, 13)
```

```{r}
primes * 2
primes - 1
```
All R statements where you create objects, assignment statements, have the same form:

object_name <- value

Use _ for spaces, object names must start with a letter and can not use numbers
# Comments
Anything after # is ignored in a line, allows someone to document work and communicate to read

# Calling Functions

```{r}
seq(from = 1, to = 10)
```
```{r}
seq(1, 10)
```

# Exercises pt 1 of 1
# Question 1
You have to put "my_variable" in an argument that would display the object like print()

# Question 2
```{r}
library(tidyverse)

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(x = displ, y = hwy)) +
  geom_smooth(method = "lm")
  
```
# Question 3
The keyboard shortcut reference commands are displayed, you can find the same information the the "Tools" tab.

# Question 4 
```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar() 

my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() 

ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```
The first plot, because its associated with the object name "my_bar_plot"
