---
title: "Chapter 18, Missing Values"
format: gfm
---

```{r}
library(tidyverse)

```

# Last Observation Carried Forward

```{r}
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4
)
```

“last observation carried forward”: You can fill in these missing values with tidyr::fill()
```{r}
treatment |>
  fill(everything())
```

# Fixed Values
use dplyr::coalesce() to replace missing values that represent some fixed and known value:
```{r}
x <- c(1, 4, 5, 7, NA)
coalesce(x, 0)
```

```{r}
x <- c(1, 4, 5, 7, -99)
na_if(x, -99)
```

# NaN
NaN (pronounced “nan”), or not a number
```{r}
x <- c(NA, NaN)
x * 10
x == 1
is.na(x)
```

If you need to distinguish an NA from a NaN, you can use is.nan(x)
```{r}
0 / 0 
0 * Inf
Inf - Inf
sqrt(-1)
```
# Implicit Missing Values
```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

# Pivoting
```{r}
stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  )
```

# Complete
complete() allows you to generate explicit missing values by providing a set of variables that define the combination of rows that should exist: 
```{r}
stocks |>
  complete(year, qtr)
```
```{r}
stocks |>
  complete(year = 2019:2021, qtr)
```

# Joins
anti_join(x, y) is a particularly useful tool here because it selects only the rows in x that don’t have a match in y
```{r}
library(nycflights13)

flights |> 
  distinct(faa = dest) |> 
  anti_join(airports)

flights |> 
  distinct(tailnum) |> 
  anti_join(planes)
```
# Exercises pt 1 of 1
```{r}
missing_planes <- flights |> 
  anti_join(planes, by = "tailnum")

missing_planes

missing_planes |> 
  count(carrier, sort = TRUE)
```
Some carriers (like MQ or US) may have more missing planes than others.

# Factors and Empty Groups
```{r}
health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)
```

```{r}
health |> count(smoker)

health |> count(smoker, .drop = FALSE)
```

```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete()

ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```
```{r}
health |> 
  group_by(smoker, .drop = FALSE) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  )
```
```{r}
# A vector containing two missing values
x1 <- c(NA, NA)
length(x1)

# A vector containing nothing
x2 <- numeric()
length(x2)
```

```{r}
health |> 
  group_by(smoker) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  ) |> 
  complete(smoker)
```

